# Linear Regression Theory I: Simple Linear Regression {#lin-t-1}


## What is Linear Regression

When we use statistical modelling in social sciences there are two main
approaches.
The more classical approach is to use modelling for estimating the effect that
one or several independent variables have on one dependent variable. Maybe we 
are interested in knowing if a higher income has an effect on life satisfaction 
and if yes, what the direction and magnitude of this effect is. Does more money
actually make you happier?

The other and more recent approach is to use modelling for making predictions
with high accuracy. Based on the relationships between many independent 
variables and one dependent variable, we try to predict the latter for actual
or hypothetical cases and their individual values for the independent variables.
This approach lies at the heart of *machine learning* and drives many of the
technologies we use on a daily basis from E-Mail spam filters to ChatGPT.

Linear regression is one of the many available modelling techniques and it can
serve both approaches lined out above. In this session we will focus on using 
linear regression for estimating an effect of interest but we will return to
prediction at a later point in this course.

But how do we know if we should choose linear regression for a specific task?
This is not easy to answer as there are many alternatives and even variations of
linear regression which may be better suited for a specific empirical problem.
As this is an introduction to modelling and time is of the essence we opted to
reduce the options and focus on two kind of models over the next weeks. Linear
regression and logistic regression. Both are comparably easy to understand and
use. Also, if we understand both of these techniques, we are in a good position
to build upon our knowledge and learn all of those more complex and specific
models that we will encounter in textbooks and scientific papers.

With the pool of options trimmed down to two, the question remains unanswered.
Should I use linear or logistic regression for my task? But now the answer is
relatively straightforward. What is the type of our dependent variable? Is it
metric? Then we use linear regression. Is it binary or categorical? Then we use
logistic regression. Linear regression will be the focus of this and the next
weeks and then we will turn to logistic regression.

Now let us dive in and learn what linear regression is all about.

## Examplary research question & data

Let us imagine that we are interested in a research question that asks what
makes a good grade in a seminar paper. In particular we are interested in the
effect that the hours a student invests in working on it has on the grade. Based
on some theoretical considerations, and maybe some idealistic views, we derive
our main hypotheses that putting in more hours will result in a better grade.

Now we also - hypothetically - held a small survey and asked 200 imaginary
students some questions on how they approached writing a seminar paper. In
particular we asked them how much time they spent working on the paper, if they
have attended (almost) all seminar sessions, how closely they worked with their
lecturers in preparing the paper and what the mean grade for previous papers
was. As these imaginary students have already turned in their papers, we also
know the grades they achieved.

Please note, that this is data on **imaginary** students, meaning we have
simulated the data making some assumptions on how to achieve a good (or bad)
grade in a paper. The assumptions we made do not necessarily reflect the way
*you* write a good paper while still being based in our experience on what it
takes to achieve a good grade. But remember, no real students were harmed in
making up this example data.

``` {r simulate_grades_data, include = FALSE}
library(tidyverse)

set.seed(12345)

grades <- tibble(
  hours = rpois(200, 40),
  previous_grades = rnorm(200, mean = 3, sd = 1),
  attendance = as.logical(rbinom(n = 200, size = 1, prob = 0.75)),
  contact = rep(c("No contact", "E-Mail", "In Person"), times = c(80, 50, 70)) %>% 
    as_factor(),
  contact_num = case_when(
    contact == "No contact" ~ -0.2,
    contact == "E-Mail" ~ 0.2,
    contact == "In Person" ~ 0.7)
) %>%
  mutate(previous_grades = case_when(
           previous_grades < 1 ~ 1.0,
           previous_grades > 5 ~ 5.0,
           TRUE ~ previous_grades) %>% 
           round(digits = 1),
         grade = previous_grades * (0.9 + rnorm(200, 0, 0.1)) -
           (hours - 40) * (0.05 + rnorm(200, 0, 0.03)) -
           attendance * (0.3 + rnorm(200, 0, 0.12)) -
           contact_num * rnorm(200, 1, 0.07) -
           (hours - 40) * (contact_num) * (0.08 + rnorm(200, 0, 0.002)) +
           0.75 - rnorm(200, 0, 0.05),
         grade = case_when(
           grade < 1 ~ 1.0,
           grade > 5 ~ 5.0,
           TRUE ~ grade
         ) %>% 
           round(digits = 1)) %>% 
  select(-contact_num) %>% 
  relocate(grade) %>% 
  arrange(grade) %>% 
  mutate(previous_grades_centered = previous_grades - mean(previous_grades)) %>% 
  mutate(hours_centered = hours - mean(hours))
  

save(grades, file = "../datasets/grades.RData")  
```


Let us have a first look on the data:
XXX ALIGN WITH EDA XXX

``` {r grades_skim, echo = FALSE}
library(skimr)
library(knitr)
kable(skim(grades))
```

Right now, the observations are ordered by the grade of the seminar paper which
run from $1.0$ to $5.0$ in increments of $0.1$. While this is somewhat
unrealistic - the german grading system actually only uses theincrements $.0$,
$.3$ and $.7$ - simulating the data in this way will make the demonstrations on
linear regression easier and more straightforward. The variable
`previous_grades` is set up in the same way and represents the mean of the
grades the student received up to this point `hours` represents the time a
student spent on writing the paper, ranging from $23 - 57$ hours, with a mean of
about $40$. Besides these metric variables, the data set also contains two 
categorical measures. `attendance` is a *dummy variable*, meaning it can only
have the values $1$ or $0$ or `TRUE` and `FALSE` in this case, as it is saved as
a logical variable. `TRUE` represents that a student attended almost all seminar
sessions before writing the paper - which about $77%$ did -, `FALSE` states that
they did not.
`contact` is a factor variable with three categories and shows the answers to
the imaginary question on how much contact the student had to the lecturer
before starting the writing process. Besides `No contact` the students could 
have had `E-Mail` contact to state ther research question and get some short
written feedback or meet the lecturer `In Person` to achieve a deeper discussion
of the question and laid out plan for writing the paper.
The two additional variables are versions of `previous_grades` and `hours` that
are centered on their respective means. They will come into play at a later 
point in this session.

Let's have a look at some observations.

``` {r grades_looksie, echo = FALSE}
grades %>% 
  head(n = 10)
```

From this first 10 rows, we can see that the students with the best grades spent
more than 40 hours on writing, have already achieved good grades in their papers
up to this point and at least had some contact to the lecturers. Most also
regularly attended the seminar but two did not and still achieved a $1.0$ in
their grade.

So what makes a bad grade?

``` {r grades_bad_looksie, echo = FALSE}
grades %>% 
  tail(n = 10)
```

Here the picture seems less clear. While most students did not put in as many
hours, some did and still failed to pass. Half of the students that received a
$5.0$ regularly attended and most at least had E-Mail contact before writing
their paper. What seems to be more consistent though is that the mean of the
previous grades is rather low.

So what do we know now? Does a good or bad track record in grades predict all
future grades? This seems not only unrealistic but also a kind of sad take home
message. To get a better understanding on which of the potential influential
variables had an effect on the final grade and what the magnitude and direction
of these effects was, we now turn to linear regression.

## Simple Linear Regression

In a *simple linear regression*, the model is used to describe the relationship
between *one* dependent and *one* independant or explanatory variable. The
question this model can answer for us is, by how much does the dependent
variable increase or decrease, when the explanatory variable increases by $1$?

Returning to our exemplary research question on what makes a good grade in a
seminar paper an intuitive hypotheses would be, that the grade gets better the
more hours a student invests in writing the paper. In this case we assume a
linear relationship between the independent variable `hours` and the dependent
variable `grade`. As german grades are better the lower their value, we thus
would assume a negative effect from `hours` on `grade`.

Before turning to the formalities and practical application of a simple linear
regression model, let us first have a look on this relationship by plotting the
variables against each other.

``` {r plot_grade_hours, echo = FALSE}
grade_vector <- c(1.0, 1.3, 1.7, 2.0, 2.3, 2.7, 3.0, 3.3, 3.7, 4.0, 4.3, 4.7, 5.0)
grade_vector_small_lbl <- c("1.0", "2.0", "3.0", "4.0", "5.0")

grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

When we are talking about dependent and independent variables, there is the
convention to plot former on the x-axis and the latter on the y-axis. So the
*y-variable* is to be explained and the *x-variable* is used to explain it.
This convention will also be used in all formulas in this seminar.

Looking at the plot we first see a cloud of dots, representing all combinations
of `hours` and `grade` in all our $200$ observations. It may be hard to pick out
any pattern, but looking closely we can observe that overall the dots seem to
follow a downward slope from the upper left - indicating few hours worked and a
worse grade - towards the lower right - indicating more invested hours and a
better grade. This would be the relationship stated in our hypotheses. The more
hours a student works on a seminar paper the better the final grade will be.

We can try to describe this pattern by adding a line from the upper left to the
lower right. 

``` {r plot_grade_hours_wild_guess, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

This describes the realtionship between the two variables as linear. Each hour
invested decreases the grade by a certain amount, for this proposed line by
exactly $0.08$ points. Remember that decreasing the value of the grade actually
means getting a better grade.

But is this the only possible line or even the *correct* one? Most certainly not
as the values used to draw the line were only a wild guess by the authors. We
could imagine several other lines that also look more or less reasonable - as
well as some that look unreasonable - and add them to the plot.

``` {r plot_grade_hours_wild_guesses, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  geom_abline(slope = -0.06, intercept = 5, colour = "blue") +
  geom_abline(slope = -0.02, intercept = 4, colour = "yellow") +
  geom_abline(slope = 0.03, intercept = 3, colour = "green") +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

While we have some intution, that especially the green line misses the mark by a
lot, we can't really decide between the others just by looking at the plot. The
data points are way to dispersed to see the relationship clearly.

The goal of using a simple linear regression model is to identify the *one* line
that describes the relationship the best. The *best* meaning, with as little
error as possible. 

XXX OVERLAPPING DOTS? XXX

### Regression Formula

To understand how these lines in the above plot were conceived and how to find 
the line with the best *fit*, i.e. the lowest error, we have to understand the
formular for linear regression. While formulas may always be kind of daunting,
we are in luck as tis particular one is actually quite easy to understand,
especially when paired with a graphical representation.

$$y = \beta_0 + \beta_1*x_1 + \epsilon$$

Let us first look at the parts we already know. $y$ is the dependent variable,
in our case the grade achieved. So one thing is for sure, the whole right part
of the equation has to be used to calculate the value of $y$ from the data, i.e.
the dependent variable $x$. Here we have three terms. Let us skip the first one
for now and focus on the second one $\beta_1*x_1$. 

$x_1$ is the dependent variable, in our case `hours`. $\beta_1$ is the 
*redression coefficient* for $x_1$. This value gives us the *slope* of the
regression line. Based on this, we can start rewriting the general formula and 
tailor it to our specific use case.

$$y_{grade} = \beta_0 + \beta_{hours}*x_{hours} + \epsilon$$


Let us return to the first wild guess we made above.

``` {r plot_grade_hours_wild_guess_expl_slope, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

Here we guessed, that an increase in time invested of one hour decreases the
value of `grade` by $0.08$. This is the slope of the red line and thus also the
coefficient in the regression formula that is used in computing said line. So,
$\beta_{hours} = -0.08$. We can insert this value into our formula.

$$y_{grade} = \beta_0 -0.08*x_{hours} + \epsilon$$

In this way the value of $x_{hours}$ is multiplied by $-0.08$. Let us assume a
student worked $40$ hours on their paper. $-0.08*40$ being $-3.2$, we know, that
working 40 hours on a paper *on average* - more on that later - leads to a $3.2$
lower grade value and thus a better grade. But $3.2$ lower than what?

Looking at the formula again, we see that subtract this value from $\beta_0$.
This is the *intercept*, the value at which the line intersects with the y-axis.
Let us zoom out on our plot, to see what happens.

``` {r plot_grade_hours_wild_guess_expl_intercept, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  scale_y_continuous(breaks = 1:6, labels = c("1.0", "2.0", "3.0", "4.0", "5.0", "6.0"), limits = c(0, 6)) +
  scale_x_continuous(breaks = seq(0, 60, 10), limits = c(0, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

We can now see the point where the red line intersects with the y-axis. This is
the intercept of this line, i.e. $\beta_0 = 6$.

$$y_{grade} = 6 -0.08*x_{hours} + \epsilon$$

If we now again assume a time investment of $40$ hours, we can compute
$6-0.08*40 = 2.8$. So our red regression line - which is still only a wild guess
- assumes, that working 40 hours on a seminar paper will result in a grade of 
$2.8$, on avarage. We can mark these values in our plot


``` {r plot_grade_hours_wild_guess_expl_40hours, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_segment(x = 0, y = 2.8, xend = 40, yend = 2.8, linetype = 2, colour = "red") +
  geom_segment(x = 40, y = 0, xend = 40, yend = 2.8, linetype = 2, colour = "red") +
  geom_point() +
  geom_point(x = 40, y = 2.8, colour = "red") +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

The red dot is the intersection of the values `hours = 40` and `grade = 2.8`.
As this is the value for $y$ our regression line assumes a student with a time
investment of 40 hours achieves, the red dot also lies exactly on the red line.

But if we look at the plot once again, we can see that most actual observations
for students that invested 40 hours do not actually lie on the regression line
but are scattered above and below the line. So some of these students achieve
much worse or much better grades than $2.8$ investing the same amount of time in
their work. This leads us to the last part of the formula, $\epsilon$.

This is the *error term*. Having data that is dispersed like this - and any real
world data will always be - our linear line will never be able to pass exactly
through every data point. Some points may lie exactly on the line, but many or
most will not.

We can visualize this. To keep the plot readable, we only do this for some 
random observations but in reality the distance of every data point from the
regression line is taken into account.

``` {r plot_grade_hours_wild_guess_expl_residuals, echo = FALSE}
set.seed(123456)

grades %>% 
  slice_sample(n = 5)

grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.08, intercept = 6, colour = "red") +
  geom_segment(x = 28, y = 6-0.08*28, xend = 28, yend = 2.3, colour = "red") +
  geom_segment(x = 39, y = 6-0.08*39, xend = 39, yend = 2.0, colour = "red") +
  geom_segment(x = 38, y = 6-0.08*38, xend = 38, yend = 3.1, colour = "red") +
  geom_segment(x = 40, y = 6-0.08*40, xend = 40, yend = 3.3, colour = "red") +
  geom_segment(x = 44, y = 6-0.08*44, xend = 44, yend = 4.5, colour = "red") +
  geom_point(x = 28, y = 2.3, colour = "red") +
  geom_point(x = 39, y = 2.0, colour = "red") +
  geom_point(x = 38, y = 3.1, colour = "red") +
  geom_point(x = 40, y = 3.3, colour = "red") +
  geom_point(x = 44, y = 4.5, colour = "red") +
  scale_y_continuous(breaks = 1:5, labels = grade_vector_small_lbl) +
  scale_x_continuous(breaks = seq(20, 60, 10), limits = c(20, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

The distance of these or rather all points from the line, the *residuals*, are
represented in the error term $\epsilon$. It is a measure for how wrong our line
is in describing the data in it's entirety. So why is it wrong? We can not say
for sure, but there are two common main reasons. 

For one, there may be other variables that also influence the relationship
between invested hours and achieved grade, something that we will return to
later in this session when we expand the idea of linear regression to multiple
independent variables. 

But there is also random variation present in every bit of real world data.
While our data is simulated we also added random variation on purpose. Because
this is what real world data is, it's messy and it's noisy.

Not every seminar paper that had the same time investment, e.g. 40 hours, will
have the same quality in results. There may be other influential variables, e.g.
the student's general skill level or if they sought assistance by their lecturer
in preparing the paper, influencing the final grade. But even if the quality of
the paper after working 40 hours would be the same for each student, measurement
error, i.e. noise, will be introduced because not every lecturer will grade
exactly the same or maybe because papers were submitted at different time
points and grading standards may have changed. If we can not measure these
variables we have to accept these unobservable sources of noise and hope, where
*hope* actually means thorough theoretical and methodical thinking, that we can
still measure our effect of interest. But his also means, that measuring and
modelling **always** includes uncertainty. We never know for certain if and to
what extent our results are influenced by unobservable variables and random
variation. Still, there are ways to assess this uncertainty, which we will
regularly return to during the course. This should not stop quantitative
social scientists from making strong or even bold arguments based in thorough
theoretical thinking and responsible data analysis, but we always have to 
acknowledge the uncertainty included in every step and make it a part of our
interpretations and conclusions.

XXX MAYBE SHORTEN THE SERMON... XXX

The error term $\epsilon$ is the final piece of the puzzle in actually computing
a linear regression model. Without jumping into the mathematics of it all, the
technique that is used to estimate the coefficients $\beta_0$ and $\beta_1$ is
called *OLS* - Ordinary Least Squares. What it basically does, is to take the
squares of all residuals, i.e. the distances of the data points from the 
regression line, sum them up and minimise this value. All this substantialy
means is, that OLS searches the regression line with lowest amount of error,
i.e. the lowest overall distance from the actual data points.

This computation gives us estimates for the regression coefficients in this
formula:

$\hat{y} = b_0 +b_1*x_1$

We can see two differences to the formula we started with. First, we write
$\hat{y}$ - pronounced as "y hat" - instead of $y$. At the same time, we exclude
the error term $\epsilon$. This means that we are no longer computing the actual
value of $y$, as in the point on the regression line for a certain value of
$x_1$ $+$ the error, but the estimate $\hat{y}$, as in the point on the
regression line that is predicted for a certain value of $x_1$. Second, we write
$b$ instead of $\beta$. This also alludes to the fact that we are now
computing an estimate for the coefficients based on the data available and not
the real and unknown value of $\beta$.

XXX MAYBE ALSO SOME CUTTING REQUIRED XXX

XXX include "on average" XXX

### Regressing `grade` on `hours`

Now that we have a firmer understanding on what linear regression actually is
and does, we can finally get to the fun part and use the technique for
estimating the effect of `hours` on `grade` or in other words, regress
`grade` on `hours`.

``` {r reg_grade_hours, echo = FALSE}
lm(grade ~ hours, grades) %>%
  summary()
```

This is the output from a simple linear regression for `grade` on `hours` from
R. How to do this in practice and what the first to lines mean will be the topic
of the next sesion. For now we will focus on the coefficient block and
introduce the additional elements of the output on by one during this session.

The column `Estimate` gives us the values for $\beta_0$ and $\beta_1$ discussed
above. The estimated coefficient for `hours` tells us that out intuition was
right, the more hours a student invests in writing a paper, the better the grade
will be. In this case every additional hour spent on working will decrease the
value of the grade by $-0.05236$ points. In keeping with the example of a 40
hour workload this leads to a decrease of $-0.05236 * 40 = -2.0944$ points.
Adding the intercept from the same column, the estimated grade after working 40
hours is $5.07912 -0.05236 * 40 = 2.98472$. So on average a student from our
simulated data set will pass after 40 hours of work but will not get a great
grade.
Remember, this is the expected average value. This does not mean that some
students will not get better or worse grades, or even fail to pass, after this
amount of time investment, as we hav seen in the plot.

Now that we know the coefficients for the regression line with the best fit,
i.e. the lowest error, we can again visualise the result.

``` {r plot_lm_grade_hours, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.05236, intercept = 5.07912, colour = "blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  scale_y_continuous(breaks = 1:6, labels = c("1.0", "2.0", "3.0", "4.0", "5.0", "6.0"), limits = c(0, 6)) +
  scale_x_continuous(breaks = seq(0, 60, 10), limits = c(0, 60)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested",
       y = "Grade achieved")
```

Now, what grade can a student expect, on average, if they invest exactly 0
hours, i.e. do nothing and hand in a blank paper. We can look at the graph or to
achieve a more precise result, calculate it.

$$5.07912 -0.05236 * 0 = 5.07912$$

As $x_{hours} = 0$ for this theoretical example, the estimated value $\hat{y}$
or $y_{grade}$ is the same as the intercept $\beta_0$. This is what the
intercept represents in general, the estimated value $\hat{y}$ when the
dependent variable is $0$.

Now, investing zero hours in a seminar paper is not only not advisable, it is
also not a value we observed in our data. If the data would include observations
with zero hours of time invested, the grade would be a firm $5.0$ and the same 
would be true for low single digits, i.e. turning in a two-pager as a seminar
paper. The takeaway is, that the model is highly dependent on the data that it
is trained on. If the data would have included such cases we could expect a
higher intercept and a steeper slope, i.e. stronger negative coefficient.

Luckily all our simulated students have put in at least some hours. But as we do
not have data for zero to $22$ hours, we can not really make reliable estimates
in this range. Because of this, it does not really make sense to enter `hours`
into the regression model as ranging from $0$ to $57$. One solution that is
often used for metric variables is to center them on their mean. This can be
achieved by simply subtracting the mean of $x$ from each individual value:
$x_i - \bar{x}$.

We can now rerun the regression.

``` {r reg_grade_hours_centered, echo = FALSE}
lm(grade ~ hours_centered, grades) %>%
  summary()
```

Comparing the results to the first model shows us, that the coefficient for
$b_{hours\_centered}$ is exactly the same as for $b_{hours}$. So the effect
of working more hours has not changed. What has changed is the value of the
intercept. This will make more sense if we again plot the regression line.

``` {r plot_lm_grade_hours_centered, echo = FALSE}
grades %>% 
  ggplot(aes(x = hours_centered, y = grade)) +
  geom_point() +
  geom_abline(slope = -0.05236, intercept = 2.9675, colour = "blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  scale_y_continuous(breaks = 1:6, labels = c("1.0", "2.0", "3.0", "4.0", "5.0", "6.0"), limits = c(0, 6)) +
  scale_x_continuous(breaks = seq(-20, 20, 10), limits = c(-20, 20)) +
  labs(title = "Regression of seminar paper grade on hours invested in writing",
       x = "Hours invested (centered)",
       y = "Grade achieved")
```

By centering the x-Variable on its mean we have changed its interpretation.
A value of `hours = 0` now stands for investing as much time as is the mean
of `hours` in the whole data set, which in this case is $40.33$ hours. Positive
values indicate that a student worked $x$ hours more, negatives indicate $-x$
hours less compared to the mean. In this way, we also moved the y-axis and thus
changed the interpretation of the intercept. Its new value of $2.9675$ now
indicates the estimate for a student who invests the mean value of `hours` in
their work, i.e. $40.33$. We will use this version of the variable for the rest
of the session.


XXX OUTRO XXX