# Linear Regression Theory II: Multiple Linear Regression {#lin-t-2}

``` {r init, include = FALSE}
library(tidyverse)

load("../datasets/grades.RData")
```


XXX INTRO XXX

## Multiple Linear Regression

Maybe explaining the grade a student receives solely based on the hours of
invested time, does not paint the whole picture. As we have alluded to, there
may be other variables that could also have an effect of the final grade.

A *simple linear regression* only allows for one independent variable. This is
why we need *multiple linear regression* if we want to start introducing 
additional variables into the model. Luckily this is easy to understand as we
already know the formula for a simple linear regression.

$$y = \beta_0 + \beta_1*x_1 + \epsilon$$

To change a simple into a multiple linear regression, we just start adding the
additional variables and their coefficients additively to the formula.

$$y = \beta_0 + \beta_1*x_1 + \beta_2*x_2 + ... + \beta_k*x_k + \epsilon$$

So to add a second variable and its coeffcient we add the term $+ \beta_2*x_2$
and so on until we added all independent varibles of interest $k$ to the model.
Everything else works exactly as described above for the simple model.

### Adding additional metric variables

We already expected that the mean of the previous grades could be a strong
predictor for future grades. We could understand these as a *proxy* variable for
the general skill level of a student. The higher the skill level, the higher
previous grades will have been.

How we can add additional variables in R code will again be a topic for the next
session, but let us look at the results of a regression of `grade` on
`hours_centered` and `previous_grades_centered`, the latter being centered on the
mean previous grade of $2.935$.

``` {r reg_grade_hours_prev, echo = FALSE}
lm(grade ~ hours_centered + previous_grades_centered, grades) %>% summary()
```

As we added a new variable, we now see three coefficients.
The intercept has not changed. It now indicates the estimated grade for a
student who invests the mean amount of hours, $40.33$, and whose previous grades
are exactly $2.935$.

The coefficient for `hours_centered` got mildly more negative, still telling us
that the value of `grade` gets lower, the more hours are invested in writing the
paper. This coefficient now gives us the effect while *controlling* for the
effect of `previous_grades_centered`. This is what multiple linear regression
does, giving us the coefficients for our variables of interest while keeping all
other independent variables at specific values. As we have centered the variable
for previous grades, the coefficient for `hours centered` gives us the effect 
when the previous grades were exactly at the mean of $2,935$.

In the same way, the coefficient for `previous_grades_centered` gives us the 
effect of previous grades when the invested hours are controlled for, in this
case when the invested hours were exactly $40.33$. The coefficient is rather
high and positive. This indicates that a student with a previous grade value
that is $1$ above the mean, is estimated to receive a new grade that is $0.9$ 
points above the intercept. This means, that the previous grade is a very strong
predictor for the new grade.

While plotting in more than two dimensions gets really hard, we can still
calculate $\hat{y}$ for certain values of both independent variables.
We already know the predicted grade for a student with mean values on both 
independent variables, as this is the intercept. To make sure that we correct,
we can calculate it again.

$$b_0 + b_{hours\_centered}*0 + b_{previous\_grades\_centered}*0 = 2.9675$$

For this case we can see, that the previous grade actually is a strong
predictor, as the previous and new grades are substantially the same.

What if a student whose previous grades were $1$ above the mean, so just below
$4.0$ but who decides to invest $10$ hours more than the mean for the new paper?

$$2.9675 - 0.056543 * 10 + 0.904079 * 1 = 3.306149$$

So the good message is, while previous grades are a strong predictor, putting in
more hours still leads to better grades.

What if a really good student decides to rely on their skill and to work less
this time?

$$2.9675 - 0.056543 * -10 + 0.904079 * -2 = 1.724772$$

While $1.7$ is still a very good grade, working 10 less hours than the mean of
students leads to a substantially worse estimate compared to the about $1.0$
received in previous grades.


### Adding dummy variables

Another variable that could be of interest in explaining the received grade,
is if a student attended most of the seminar sessions.
`attendance` holds this information in the form of a dummy variable. Dummies can
only have two states. "Yes" or "No", "1" or "0" or in this case "`TRUE`" or 
"`FALSE`".

Let us add the variable to our model.

``` {r reg_grade_hours_prev_att, echo = FALSE}
lm(grade ~ hours_centered + previous_grades_centered + attendance, grades) %>% summary()
```

This gives us a new line in the R Output holding an estimate for
`attendanceTRUE`. What is meant by this? In contrast to the metric variables we 
have uses in our model up to this point, a dummy variable - or binary variable -
can only have two states. As we are using a logical variable here, it can only
have the value `TRUE` - here indicating regular attendance - or `FALSE`. So what
the output shows us, is the effect of attendance being `TRUE` compared to being
`FALSE`. If a student did regularly attend the seminar, the estimated grade is
$-0.248250$ lower compared to when they did not.

We can observe what happens in the formula:

$$\hat{y} = b_0 + b_{hours\_centered}*x_{hours\_centered} + b_{previous\_grades\_centeterd}*x_{previous\_grades\_centeterd} + b_{attendance} * x_{attendance}$$

If you calculate with `TRUE` and `FALSE` in R, the values $1$ and $0$ are used
respectively. So $x_{attendance}$ can either have the value $1$ for regular
attendance or $0$ for not so regular attendance.

If a student did regularly attend, the coefficient $b_{attendance} becomes a
part of the estimate $\hat{y}$:

$$\hat{y} = b_0 + b_{hours\_centered}*x_{hours\_centered} + b_{previous\_grades\_centeterd}*x_{previous\_grades\_centeterd} + b_{attendance} * 1$$

If stduent did not regularly attended, this happens: 

$$\hat{y} = b_0 + b_{hours\_centered}*x_{hours\_centered} + b_{previous\_grades\_centeterd}*x_{previous\_grades\_centeterd} + b_{attendance} * 0$$
$$\hat{y} = b_0 + b_{hours\_centered}*x_{hours\_centered} + b_{previous\_grades\_centeterd}*x_{previous\_grades\_centeterd}$$

The coeeffcient is no longer a part of the estimate. One can basically say, the
coefficient gets switched on or off by the value of the dummy variable.

So while the estimate for a student with mean values for invested hours and
previous grades who did not attend is the intercept of $3.157411$ for the same
student with attendance we can calculate the estimate as:

$$3.157411 - 0.053942*0 + 0.911802*0 - 0.248250 * 1 = 3.157411 - 0.248250 = 2.909161$$

It seems attending class is an easy way to raise one's grades.

### Adding categorical variables

We have one further variable in our simulated data set that could be of interest
in explaining, what makes a good grade in a seminar paper.`contact` is a factor
variable. It can take three different categories. `No contact` indicates that
the student did not contact the lecturer to discuss a research question or the
laid out plan for the paper. `E-Mail` means that there was some written contact
and at least the basics for the paper were discussed before writing. Lastly,
`In Person` stands for an in depth discussion with the lecturer, clearing up
problems beforehand and having a more stringent vision for the paper before 
writing the first word.

Let us add the variable to our model.

``` {r reg_grade_hours_prev_att_contact, echo = FALSE}
lm(grade ~ hours_centered + previous_grades_centered + attendance + contact, grades) %>% summary()
```

Wait, we entered three categories into the model and got estimates for two of
them. What happened? What R does is to create two dummy variables on the fly.
The first discerns between having E-Mail contact or no contact at all. The 
second one between having contact in person to no contact at all. So for
categorical variables in regression models we always compare being in one
category to being in the *base category*. In this case the base category is
`No contact` but we could also change the base category. It depends on what we
are interested in comparing to. For our example comparing the effects of having 
more in depth contact to having none makes sense.

Let us look at our formula again:

$$\hat{y} = b_0 + b_{hours\_centered}*x_{hours\_centered} + b_{previous\_grades\_centeterd}*x_{previous\_grades\_centeterd} + b_{attendance} * x_{attendance} + b_{E-Mail} * x_{E-Mail} + b_{In Person} * x_{In Person}$$

Now there are three possibilities. A student can have no contact at all. In this 
case both dummy variables equal $0$. To make our formula easier to read, we have
abbreviated the middle part for now:

$$\hat{y} = b_0 + ... + b_{E-Mail} * 0 + b_{In Person} * 0$$
So in this case, controlling all other independent variables at their default
values, the mean for the metric variables, `FALSE` for `attendance`, the 
intercept gives us the estimate for the grade as both dummy variables that
were created for `contact`are "switched off".

THe two other possibilities are that a student either had E-Mail contact or an
in person discussion:

$$\hat{y} = b_0 + ... + b_{E-Mail} * 1 + b_{In Person} * 0$$

$$\hat{y} = b_0 + ... + b_{E-Mail} * 0 + b_{In Person} * 1$$

In both cases the relevant dummy variable is "switched on" while the other does
not factor into the equation.

Looking at the estimates, we can see that having contact to the lecturer before
writing has strong negative effects, resulting in better grades. Having E-Mail
contact reduces the value of `grade` by $-0.413808$ points, having an in person
discussion by $-0.853252$.

So what grade can a student whose previous grades were at the mean of $2.935$,
but who decided to put in 20 hours more compared to their peers, regularly
attend the seminar and have an in-depth personal discussion before writing their
paper expect on average as their new grade?

$$3.617949 - 0.050830 * 20 + 0.874123 * 0 - 0.324653 * 1 - 0.413808 * 0 - 0.853252 * 1 = 1.423444$$

Putting in the hours, attending and working with your lecturer seems to pay off,
at least in our simulated data set.

<!-- ### Adding interactions -->

<!-- CUT FOR NOW -->

<!-- ``` {r reg_grade_hours_prev_att_contact_interaction, echo = FALSE} -->
<!-- lm(grade ~ hours_centered + previous_grades_centered + attendance + contact + hours_centered * contact, grades) %>% summary() -->
<!-- ``` -->

## Returning to our research question

Our exemplary research question concerned itself with what makes a good grade in
a seminar paper. In particular we were interested in the effect of the invested 
hours, as our main hypothesis was that more hours lead to better grades. What do
we know now? 

All analysis point towards a clear effect from `hours` on `grade`. This effect
was consistently visible in all of our models. But did we correctly identify
and estimate the effect of interest? Maybe. The problem is, we actually did not
approach the analysis correctly. In a real analysis we should **absolutely**
refrain from adding variables to our model that *could be* relevant until we are
satisfied or all available variables are bunched into one huge model. It was
fine to do this in this introduction to linear regression and for learning how
different types of variables can be used in a regression model. But in a real
project, we have to invest time to think about which variables to add because we
assume that they have a relevant effect based on theoretical assumptions about
the processes we are interested in. 

So let us do this now and vow do make this our first step in all future
endeavors. While we do not have a clear theoretical basis, we can make clear
assumptions on the data generating process and draw these in a DAG.

``` {r dag, echo = FALSE}
library(dagitty)

dagitty('dag{
        Hours [pos = "1,0"]
        Prev_Grades [pos = "1, -2"]
        Skill_Level [pos = "2, -2"]
        Attendance [pos = "2, -3"]
        Contact [pos = " 2, -1"]
        Grade [pos = "3, 0"]
        
        Hours -> Grade
        Prev_Grades -> Skill_Level -> Grade
        Attendance -> Grade
        Hours <- Contact -> Grade
}') %>% 
  plot()
```

XXX MAYBE NICEN IT UP XXX

Our central assumption, and the effect we want to identify and estimate, is the
direct effect from `hours` on `grade` in the bottom line. The more hours a 
student invests, the better the grade should be. This is our variable of 
interest and it has to be included in the model.

The assumed effect of `contact` is more complex. For one we assume that a more
in-depth contact with the lecturer will increase the grade directly. The 
research question will be more focused, the student will know what is important
to a certain lecturer, common mistakes can be avoided if they are cleared up
beforehand and so on. But we will also assume that `contact` will have an effect 
on `hours` in the sense that the hours invested can be used more efficiently if
an in-depth discussion has taken place. Instead of wasting time running into
problems that could have been avoided most of the invested time can actually go
into constructive work. This makes `contact` a confounder for `grade` and 
`hours`. Tapping into the knowledge from the last session, now we know that we
also have to control for `contact` to measure the effect of `hours` correctly.

A student's skill level will also have a direct effect on `grade`. As we do not
have a direct measure of skill in our data, we use `previous_grades` as a proxy
for skill level. `attendance` also has a direct effect on `grade` as students
who were present in the seminar will not only have learned the seminar's
contents, but will also have a better understanding of what is expected in their
seminar papers. As neither skill level - or `previous_grades` - nor `attendance`
has any connection to `hours` we also should not control for them in our model.

That leaves us with `hours` and `contact` to be included in our linear
regression, if our goal is to accuarately identify and estimate the effect of
invested time on the final grade. So let us do this:

``` {r reg_grade_hours_contact, echo = FALSE}
lm(grade ~ hours_centered + contact, grades) %>% summary()
```

This is our estimate. Each hour invested beyond the mean of $40.33$ hours
changes the grade by about $-0.05$ points. This supports our hypotheses and we
can conclude, that investing more hours into writing a seminar paper actually
is a worthwhile investment.

But remember: This is correct as long as our DAG is drawn correctly. This is
always debatable. Maybe we should assume an effect from skill level on `hours`.
The higher the skill level the more efficiently the available time can be used.
For this example we know the DAG is correct, because we have simulated the data
exactly in this way. For real world applications we never know if our DAG is
correct. All we can - and have to - do is base it on thorough thought, 
theoretical work and sound arguments. 

This all is true *if* our goal is to estimate an effect of interest as precisely
as possible. But as we have alluded to in the introduction to this session we
could also use modelling with a different goal, i.e. predicting a grade as
accurately as possible. For this task, the model which only includes `hours` and
`contact` will not do the best job. From our DAG we know that `attendance` and
`previous_grade` should have an effect on `grade`, as we have also seen in our 
models. For this task the full model including all these variables will produce
better estimates. We will return to this in a later session, but for now we
should remember that we have to know our task because the task dictates which is
the best model to use.


## Interpretation

XXX WHERE TO PUT IT? XXX

Looking at the coefficient block from either of the two outputs, we see more
than just our estimate. The `Std. Error` or `std.error` is a measure for the
uncertainty of our estimates. It basically tells us, how far away the actual 
values of the observations used to compute the model are from our estimate
*on average*. The smaller the *standard error*, the more accurate is our
estimate. The standard error is presented in the units of the estimate and we
can thus compare them. A large standard error for a large estimate is far less
problematic compared to a large standard error for a small estimate.

The estimate and it's standard error are the basis for *hypothesis testing*.
What we are testing is the *alternative hypotheses* $H_a$ that there actually
is an effect of our independent variable on the dependent variable against the
*null hypothesis* $H_0$ that there is no effect. To reject the null hypothesis
and be confident that we are observing an actual effect, versus an effect that
is joust based on random variation in our sample, the estimate has to be far
away enough from $0$ and be accurate enough, i.e. have a small standard error.
This relationship is computed in the *t-statistic*, `t value` and `statistic` in
our outputs. From this the *p-value* can be computed, `Pr(>|t|)` and `p.value`
in the outputs. The *p-value* tells us the probability to observe a association
between the independent and the dependent variable as large or larger than our
estimate suggests, if the true association would actually be $0$. If the p-value
is small enough, we can reject $H_0$ and conclude that we observed an actual 
effect. There are certain agreed upon cutoffs in statistics where values that
meet this cutoff are considered *statistically significant*. The most common
cutoff in social sciences is $0.05$ indicated by one `*` in the output from
`summary()` There are other common cutoffs indicated by more asterisks.

Interpreting p-values correctly and not falling into the common pitfalls is a
topic on its own. We do not have the time to dive into this here, so for now we
can agree that p-values below $0.05$ indicate that we can reject $H_0$ and thus
conclude that we have actually observed an effect. Still, our interpretation of
regression results should not focus solely on p-values or lead us to disregard
any effects that did not meet the cutoff. For example, we can have very small
p-values for effects that are so small that they are substantially irrelevant.
One way to address this is to inspect the actual magnitudes of the effects,
something we will return to in our session on prediction. On the other hand, we
can have p-values larger than $0.05$ for effects that are still relevant. Maybe
the problem is not that there is no effect but that we were not able to measure
the variable in question precisely enough or that we just did not have enough 
observations. We can not go any deeper than this here, but we should remember
that the practice of declaring every effect with stars a win and disregarding
everything without them may still be common but is not the way to go forward.

XXX CORRECT? MAYBE SHORTEN XXX


XXX OUTRO XXX